Supplementary Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-02-20 paper.tex
start formatting changes for Nature

\usepackage{lineno}
\linenumbers
\renewcommand{\Affilfont}{\normalsize}
\renewcommand{\Authfont}{\normalsize}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/proposed-models.png}
    \caption{
        \textbf{Proposed conceptual models of early human history in Africa.}
        These models have been designed to translate models from the 
        paleoanthropological literature into genetically testable demographic
        models \citep{Henn2018-rf}.
        We used these conceptual models as starting points to build detailed
        parameterized demographic models (SI Section~\ref{sec:modelspec})
        that were then fit to genetic data. 
    }
    \label{fig:proposed-models}
\end{figure}

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierra Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendants of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged $550$ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/samples-diversity.pdf}
    \caption[width=\textwidth]{
        \textbf{Genetic diversity across Africa.}
        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region (bold labels)
        to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
    }
    \label{fig:diversity}
\end{figure}

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an African archaic hominin lineage or population
structure prior to the expansion (Figure~\ref{fig:proposed-models}A or
Figure~\ref{fig:proposed-models}D). As expected \citep{Ragsdale2019-nt}, this
first model was a poor fit to the data qualitatively
(Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
($LL$) $\approx-$189,300, Table~\ref{tab:supp-single-origin}). We next
explored a suite of parameterized models in which population structure predates
the differentiation of contemporary groups (Supporting Information (SI)
Section~\ref{sec:modelspec}). Depending on the parameters, these encompassed
models allowing for ancestral reticulation, such as
fragmentation-and-coalescence or meta-population models
(Figure~\ref{fig:proposed-models}B), archaic hominin admixture
(Figure~\ref{fig:proposed-models}C), and African multi-regionalism
(Figure~\ref{fig:proposed-models}D). The recent expansion and the African
multi-regional models (Figure~\ref{fig:proposed-models}A and D) have the same
topology, so interpretation of the model depends on the specified or inferred
divergence times.

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.65\textwidth]{figures/best-fit-models}
    \caption{
        \textbf{A weakly structured stem best describes two-locus statistics.}
        In the two best-fitting parameterizations of early population structure,
        continuous migration (A) and multiple mergers (B), models
        that include ongoing migration between stem populations outperform
        those in which stem populations are isolated. Most recent populations are
        connected by continuous, reciprocal migration that are indicated by 
        double-headed arrows (labels matched to migration rates and divergence
        times in Table~\ref{tab:migration-rates}). These migrations last for the
        duration of co-existence of contemporaneous populations with constant
        migration rates over those intervals. The
        merger-with-stem-migration model (B, with  $LL=-101,600$) outperformed the
        continuous-migration model (A, with $LL=-115,300$).
        Colors are used to distinguish overlapping branches and link to
        Figure~\ref{fig:diversity}.}
    \label{fig:best-fit-models}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see SI Section~S\ref{sec:f4}
        for details and additional comparisons in
        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last $100$ka.
    }
    \label{fig:predictions}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/validation}
    \caption{
        \textbf{Model validation using independent statistics.} (A--C) Using
        our best fit models to linkage disequilibrium and pairwise diversity statistics,
        we simulated expected conditional site-frequency-spectra (cSFS) and
        compared to the observed cSFS from the data. Our inferred models
        provide a good fit to the data, even though these were not used in our
        inference. Across the three populations, ancestral state
        misidentification was consistently inferred to be $1.5-1.7\%$ for
        intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
        which we used to estimate coalescence rate trajectories
        and cross-coalescence rates between pairs of populations. While
        coalescence rate distributions are informative statistics about past
        evolutionary processes, interpretation can be hindered by migration and
        population structure, and translating relative cross-coalescence rate
        curves (RCCR) into population divergence times is especially prone
        to misinterpretation. For example, the Mende-Gumuz comparison shows
        a more recent increased RCCR than either population with the British,
        a pattern that is recapitulated under our best-fit model, even though
        the Mende-Gumuz split occurs prior to the Gumuz-British split.
    }
    \label{fig:validation}
\end{figure}

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred
models. Many previous studies have found a reduction of coalescence rates
between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
could be due to either an increase in population size or to ancestral
population structure during the Middle Pleistocene \citep{Mazet2016-wn}.  All
models, including the single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between $100$ka--$1$Ma
%extending deeper in the past than the Relate estimate
% APR: I don't understand this phrase I commented out
(Figure~\ref{fig:supp-iicr-sim} and SI
Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
increase in $N_e$ during that period, the best-fit model recapitulate this
pattern without corresponding population size changes.

@@

\subsection*{The Formation of Population Structure in Africa}

@@

@@

\bibliographystyle{naturemag}
\bibliography{paper}
\end{document}

>>>>

\usepackage{lineno}
%\linenumbers
\renewcommand{\Affilfont}{\normalsize}
\renewcommand{\Authfont}{\normalsize}

@@

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierra Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendants of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}).
\aprcomment{See Supplementary Information (SI) section 1.2 for details of
sample collection and consent.}
The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged $550$ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an African archaic hominin lineage or population
structure prior to the expansion (Figure~\ref{fig:proposed-models}A or
Figure~\ref{fig:proposed-models}D). As expected \citep{Ragsdale2019-nt}, this
first model was a poor fit to the data qualitatively
(Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
($LL$) $\approx-$189,300, Table~\ref{tab:supp-single-origin}). We next
explored a suite of parameterized models in which population structure predates
the differentiation of contemporary groups (SI
Section~\ref{sec:modelspec}). Depending on the parameters, these encompassed
models allowing for ancestral reticulation, such as
fragmentation-and-coalescence or meta-population models
(Figure~\ref{fig:proposed-models}B), archaic hominin admixture
(Figure~\ref{fig:proposed-models}C), and African multi-regionalism
(Figure~\ref{fig:proposed-models}D). The recent expansion and the African
multi-regional models (Figure~\ref{fig:proposed-models}A and D) have the same
topology, so interpretation of the model depends on the specified or inferred
divergence times.

@@

@@

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred
models. Many previous studies have found a reduction of coalescence rates
between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
could be due to either an increase in population size or to ancestral
population structure during the Middle Pleistocene \citep{Mazet2016-wn}.  All
models, including the single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between $100$ka--$1$Ma
(Figure~\ref{fig:supp-iicr-sim} and SI
Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
increase in $N_e$ during that period, the best-fit model recapitulate this
pattern without corresponding population size changes.

@@

\subsection*{The formation of population structure in Africa}

@@

\bibliographystyle{naturemag}
\bibliography{paper}

\section*{Supplementary information}
Available in the online version of the paper.

\section*{Competing interests}
The authors declare no competing financial interests.

\section*{Author contributions}
\aprcomment{To do: fill in (e.g., designed the study, collected data,
analyzed data, wrote and edited the manuscript)}
\begin{itemize}
    \item APR:
    \item TDW:
    \item EGA: 
    \item EH:
    \item MM:
    \item BMH:
    \item SG:
\end{itemize}

\section*{Methods}

\aprcomment{This can be very short}

@@

\break

\section*{Figure captions}

\begin{figure}[ht]
    \centering
    %\includegraphics[width=0.5\textwidth]{figures/proposed-models.png}
    \caption{
        \textbf{Proposed conceptual models of early human history in Africa.}
        These models have been designed to translate models from the 
        paleoanthropological literature into genetically testable demographic
        models \citep{Henn2018-rf}.
        We used these conceptual models as starting points to build detailed
        parameterized demographic models (SI Section~\ref{sec:modelspec})
        that were then fit to genetic data. 
    }
    \label{fig:proposed-models}
\end{figure}

\begin{figure}[ht!]
    \centering
    %\includegraphics{figures/samples-diversity.pdf}
    \caption[width=\textwidth]{
        \textbf{Genetic diversity across Africa.}
        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region (bold labels)
        to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
    }
    \label{fig:diversity}
\end{figure}

\begin{figure}[t!]
    \centering
    %\includegraphics[width=0.65\textwidth]{figures/best-fit-models}
    \caption{
        \textbf{A weakly structured stem best describes two-locus statistics.}
        In the two best-fitting parameterizations of early population structure,
        continuous migration (A) and multiple mergers (B), models
        that include ongoing migration between stem populations outperform
        those in which stem populations are isolated. Most recent populations are
        connected by continuous, reciprocal migration that are indicated by 
        double-headed arrows (labels matched to migration rates and divergence
        times in Table~\ref{tab:migration-rates}). These migrations last for the
        duration of co-existence of contemporaneous populations with constant
        migration rates over those intervals. The
        merger-with-stem-migration model (B, with  $LL=-101,600$) outperformed the
        continuous-migration model (A, with $LL=-115,300$).
        Colors are used to distinguish overlapping branches and link to
        Figure~\ref{fig:diversity}.}
    \label{fig:best-fit-models}
\end{figure}

\begin{figure}[t!]
    \centering
    %\includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see SI Section~S\ref{sec:f4}
        for details and additional comparisons in
        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last $100$ka.
    }
    \label{fig:predictions}
\end{figure}

\begin{figure}[t!]
    \centering
    %\includegraphics{figures/validation}
    \caption{
        \textbf{Model validation using independent statistics.} (A--C) Using
        our best fit models to linkage disequilibrium and pairwise diversity statistics,
        we simulated expected conditional site-frequency-spectra (cSFS) and
        compared to the observed cSFS from the data. Our inferred models
        provide a good fit to the data, even though these were not used in our
        inference. Across the three populations, ancestral state
        misidentification was consistently inferred to be $1.5-1.7\%$ for
        intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
        which we used to estimate coalescence rate trajectories
        and cross-coalescence rates between pairs of populations. While
        coalescence rate distributions are informative statistics about past
        evolutionary processes, interpretation can be hindered by migration and
        population structure, and translating relative cross-coalescence rate
        curves (RCCR) into population divergence times is especially prone
        to misinterpretation. For example, the Mende-Gumuz comparison shows
        a more recent increased RCCR than either population with the British,
        a pattern that is recapitulated under our best-fit model, even though
        the Mende-Gumuz split occurs prior to the Gumuz-British split.
    }
    \label{fig:validation}
\end{figure}

\end{document}

### 2023-02-20 supplement.tex
start formatting changes for Nature

\begin{document}
\title{Supporting Information for\\
``A weakly structured stem for human origins in Africa''}
\author{Aaron P. Ragsdale, Timothy D. Weaver, Elizabeth G. Atkinson, Eileen Hoal, Marlo M\"{o}ller, Brenna M. Henn, and Simon Gravel}
\date{\normalsize \today}
\maketitle

@@

\addcontentsline{toc}{section}{Supporting tables and figures}
\section*{Supporting tables and figures}

>>>>

\begin{document}
\title{Supplementary Information for\\
``A weakly structured stem for human origins in Africa''}
\author{Aaron P. Ragsdale, Timothy D. Weaver, Elizabeth G. Atkinson, Eileen Hoal, Marlo M\"{o}ller, Brenna M. Henn, and Simon Gravel}
\date{\normalsize \today}
\maketitle

@@

\addcontentsline{toc}{section}{Supplementary tables and figures}
\section*{Supplementary tables and figures}
